A weakly structured stem for human origins in Africa

### 2022-01-18 paper.tex
some minor edits to discussion

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly structured stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar. 
Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stems.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human \bmhcomment{we use
this phrase a lot. human or hominin? Tim?} population structure, we sought to
capture model uncertainty by exploring multiple parameterizations of early
human population structure. The best-fit models presented here include
reticulation and low-level migration between early human populations rather
than archaic admixture. We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches (and therefore additional
parameters) during that period. \sgcomment{Some statement about hitting the
boundary} Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent, e.g., with a coalescence and
fragmentation model \citep{Scerri2019-xg}.  Eventual ancient DNA samples from
Africa could further distinguish the archaic admixture model from the weakly
structured stem model. \sgcomment{Or models involving more diversity within
Africa?}

@@

By contrast, our inferred models paint a more consistent picture of the late
Middle Stone Age as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(Supplement). During the Middle Stone Age, the Multiple Merger indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with this
period of divergence, isolation and then a merger event (i.e. a “reticulation”)
out-performed models with bifurcating divergence and continuous gene flow.

@@

Evidence for archaic admixture in Eurasia has bolstered the plausibility of
archaic admixture having also occurred in Africa. For this reason, previous
work has focused on archaic admixture to explain patterns of polymorphism
inconsistent with a single origin model. Here, we have shown that weakly
structured stem models better capture these patterns. They also help explain an
ecological riddle posed by the archaic admixture model. Neanderthal populations
were separated from early \emph{Homo sapiens} by thousands of kilometers and
continental geographic barriers. By contrast, an archaic hominin population in
Africa would need to have stayed in relative reproductive isolation from the
human lineage over hundreds of thousands of years despite closer geographic
proximity and reproductive compatibility. The weakly structured stem model
resolves this ecological riddle by accommodating continuous or recurrent
contact between two or more groups present in Africa.

There is evidence for both deleterious and adaptive archaic-derived alleles in
modern genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-related
haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
\citet{Zhang2021-xx}).\sgcomment{Probably not the best reference} Under an
archaic African admixture model, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-derived haplotypes over a
short period \citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly structured stem model in Africa.

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
assume that this correspondence also holds for early human ancestral
populations in Africa, the weakly structured stem model would predict that the
fossils deriving from each of the two stems should be morphologically quite
similar (supplementary text X)\aprcomment{Will we still include this?}, even if
they are from different parts of Africa, with differences across stems closer
to those across contemporary human populations than to human-Neanderthal
differences. Differences within stems would be comparable to differences within
recent human populations. Identifying the morphological signature of ancient
structure in the fossil record will require careful analysis of many broadly
similar remains. Our model further predicts that morphologically distinct
crania such as XX and YY are unlikely representatives of branches that
contributed appreciably to modern human ancestries.\sgcomment{Here I am being
somewhat forceful. We could use conditional to emphasize that these conclusion
are under the weakly structured stem model.}

@@

See Supplement, for now.

>>>>

\abstract{
While a modern human origin within Africa is now broadly accepted, considerable
uncertainty surrounds specific models of divergence and migration across the
continent. Progress is hampered by a paucity of fossil and genomic data, as
well as variability in dating. Here we use linkage disequilibrium and
diversity-based statistics, optimized for rapid, complex demographic inference
to discriminate among such models. We infer detailed demographic models for
populations across Africa, including representatives from eastern and western
groups, as well as 44 newly whole-genome sequenced individuals from the Nama
(Khoe-San). Despite the complexity of African population history, present-day
population structure dates back to Marine Isotope Stage (MIS) 5. The earliest
population divergence among contemporary populations occurs 120-135kya, between
the Khoe-San and other groups. Prior to the divergence of contemporary African
groups, we infer long-lasting structure between two or more weakly
differentiated ancestral Homo spp. populations connected by gene flow over
hundreds of thousands of years (i.e. a weakly structured stem). We find that
weakly-structured-stem models provide more likely explanations of polymorphism
that had previously been attributed to contributions from archaic hominins in
Africa.
In contrast to models with archaic introgression, we predict that fossil
remains from coexisting ancestral populations should be morphologically
similar. 
Despite genetic similarity between these populations, an inferred \sgcomment{4 to 10\%} of random genetic differences among [modern human populations] can be attributed to genetic drift between stems.
We show that model misspecification explains variation in previous
divergence time estimates and argue that studying a suite of models is key to
robust inferences about deep history.
}

@@

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single origin model must be explained by archaic
admixture. This is also true of this study. While it remains prohibitive to
fully explore the space of plausible models of early human population structure,
we sought to capture model uncertainty by exploring multiple parameterizations of
early history.
\bmhcomment{we use this phrase a lot. human or hominin? Tim?}
The best-fit models presented here include reticulation and migration between
early human populations rather than archaic admixture from long-isolated branches.
We cannot rule out that more complex models involving
additional stems, or hybrid models including both weak structure and archaic
admixture may better explain the data. Because parameters related to the split
time, migration rates, and relative sizes of the early stems were variable
across models, reflecting a degree of confounding among these parameters, we
refrained from introducing additional branches (and therefore additional
parameters) during that period.
\sgcomment{Some statement about hitting the boundary}
Rather than interpreting the two stems as representing well-defined
and stable populations over hundreds of thousands of years, we interpret the
weakly structured stem as consistent, e.g., with a coalescence and
fragmentation model \citep{Scerri2019-xg}.
Models including additional diversity within Africa,
and eventual ancient DNA samples from Africa, could further distinguish
the archaic admixture model from the weakly-structured-stem model.

@@

By contrast, our inferred models paint a more consistent picture of the late
Middle Stone Age as a critical period of change, assuming that estimates from
the recombination clock accurately relate to geological chronologies
(Supporting Information).
During the Middle Stone Age, the Multiple Merger indicates three
major stem lineages in Africa, tentatively assigned to southern (Stem 1S),
eastern (Stem 1E) and western/central Africa (Stem 2). While the length of
isolation among the stems is variable across model fits, models with this
period of divergence, isolation and then a merger event (i.e. a “reticulation”)
out-performed models with bifurcating divergence and continuous gene flow.

@@

Evidence for archaic admixture in Eurasia has bolstered the plausibility of
archaic admixture having also occurred in Africa. For this reason, previous
work has focused on archaic admixture to explain patterns of polymorphism
inconsistent with a single origin model. Here, we have shown that
weakly-structured-stem models better capture these patterns.
They also help explain an
ecological riddle posed by the archaic admixture model. Neanderthal populations
were separated from early \emph{Homo sapiens} by thousands of kilometers and
continental geographic barriers. By contrast, an archaic hominin population in
Africa would need to have stayed in relative reproductive isolation from the
human lineage over hundreds of thousands of years despite closer geographic
proximity and reproductive compatibility. The weakly-structured-stem model
resolves this ecological riddle by accommodating continuous or recurrent
contact between two or more groups present in Africa.

There is evidence for both deleterious and adaptive archaic-derived alleles in
modern genomes in the form of a depletion of Neanderthal ancestry in regulatory
regions \citep{Petr2019-xo} or an increased frequency of archaic-related
haplotypes such as at \emph{EPAS1} among Tibetans (e.g.,
\citet{Zhang2021-xx}).\sgcomment{Probably not the best reference} Under an
archaic African admixture model, the estimated 8--10\% introgression rate is
much higher than Neanderthal gene flow, and would have plausibly been fertile
ground for dramatic selection for or against archaic-derived haplotypes over a
short period \citep{Wall2019-ao}. By contrast, adaptation under a weakly
structured stem would have occurred continuously over much longer periods.
Patterns of polymorphism that are inconsistent with the single-stem model
predictions have been used to infer putative archaic admixed segments
\citep{Plagnol2006-lt,Hsieh2016-gk,Wall2019-ao,Durvasula2020-td}, negative
selection against such segments \citep{Wall2019-ao}, and pervasive positive
selection \citep{Schrider2017-kl}. However, such approaches are subject to high
false positives in the presence of population structure with migration
\citep{Petr2019-xo}, and their interpretation should be re-examined in light of
a weakly-structured-stem model in Africa.

Multiple studies have shown a correspondence between phenotypic
differentiation, usually assessed with measurements of the cranium, and genetic
differentiation among human populations and between humans and Neanderthals
\citep{Relethford1994-mh,Weaver2008-ho,Von_Cramon-Taubadel2009-zb}. If we
assume that this correspondence also holds for early human ancestral
populations in Africa, the weakly-structured-stem model would predict that the
fossils deriving from each of the two stems should be morphologically quite
similar (supplementary text X)\aprcomment{Will we still include this?}, even if
they are from different parts of Africa, with differences across stems closer
to those across contemporary human populations than to human-Neanderthal
differences. Differences within stems would be comparable to differences within
recent human populations. Identifying the morphological signature of ancient
structure in the fossil record will require careful analysis of many broadly
similar remains. Our model further predicts that morphologically distinct
crania such as XX and YY are unlikely representatives of branches that
contributed appreciably to modern human ancestries.\sgcomment{Here I am being
somewhat forceful. We could use conditional to emphasize that these conclusion
are under the weakly structured stem model.}

@@

Methods are detailed in the Supporting Information.
